---
title: "Data Visualization with R"
output: 
  learnr::tutorial:
    progressive: false #true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
library(learnr)
library(tidyverse)
library(gradethis)
#gradethis_setup()

knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE,
                      fig.height = 3,
                      fig.width = 5,
                      fig.align = "center")

# prepare dataset
comics <- read_csv("data/comics.csv")
#comics <- read_csv("data/comics.csv") %>% 
#  filter(!is.na(align), 
#         !is.na(id), 
#         !is.na(gender)) %>%
#  droplevels()


```

## Overview

Visualization is one of the most important tools for data science.

The tutorial focuses on three basic skills:

1.  How to create graphs with a reusable **template**
2.  How to add variables to a graph with **aesthetics**
3.  How to make different "types" of graphs with **geoms**

### Loading the Tidyverse package

Packages are the fundamental units of reproducible R code. They include reusable functions, the documentation that describes how to use them, and sample data. In this lesson we will make use of two packages:

Tidyverse is a collection of R packages for data science that adhere to a common philosophy of data and R programming syntax, and are designed to work together naturally. You can learn more about tidyverse [here](https://tidyverse.org/). But no need to go digging through the package documentation, we will walk you through what you need to know about these packages as they become relevant.

Once we have installed the packages, we use the `library()` function to load packages into R.

Let's load the packages to be used in the remainder of this lesson.

```{r load-packages, exercise=TRUE}
library(tidyverse)
```



## The mpg dataset


The `mpg` sample dataset is included in the `ggplot2` package. `mpg` contains observations collected on 38 models of cars by the US Environmental Protection Agency.

To see the `mpg` data frame, type `mpg` in the code block below and click "Submit Answer". 
You can use the black triangle that appears at the top right of the table to scroll through all of the columns in `mpg`.

```{r mpg0, exercise = TRUE}

```

```{r mpg0-solution}
mpg
```


```{r mpg0-check}
grade_this({
  print(.result)
  print(.solution)
  # custom checking code appears here
  if (identical(.result, .solution)) {
    pass("Good job!  We'll use interactive code chunks like this throughout these tutorials. Whenever you encounter one, you can click Submit Answer to run (or re-run) the code in the chunk. If there is a Solution button, you can click it to see the answer.")
  }
  
  # call log for mistake
  fail("Please try again!")
})

#grade_code()
```

## Scatterplots

### 

Among the variables in `mpg` are:

1.  `displ`, a car's engine size, in liters.
2.  `hwy`, a car's fuel efficiency on the highway, in miles per gallon (mpg). A car with a low mpg consumes more fuel than a car with a high mpg when they travel the same distance.

Now let's use this data to make our first graph.

### Creating a scatterplot

The code below uses functions from the **ggplot2** package to plot the relationship between `displ` and `hwy`.

To see the plot, click "Run Code."

```{r ggplot1, exercise = TRUE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

Can you spot the relationship?


```{r mc, echo=FALSE}
question("Which of the following statement correctly describes the relationship between between engine size (`displ`) and fuel efficiency (`hwy`)?",
  answer("The plot shows a positive relationship between engine size (`displ`) and fuel efficiency (`hwy`). The cars with big engines use *more* fuel."),
  answer("The plot shows a positive relationship between engine size (`displ`) and fuel efficiency (`hwy`). The cars with big engines use *less* fuel.",  message = "Try again!"), 
  answer("The plot shows a negative relationship between engine size (`displ`) and fuel efficiency (`hwy`). The cars with big engines use *more* fuel.", correct = TRUE),
  answer("The plot shows a negative relationship between engine size (`displ`) and fuel efficiency (`hwy`). The cars with big engines use *less* fuel.",  message = "Try again!"), 
  allow_retry = TRUE
)
```



### A closer look

In the plot below, one group of points (highlighted in red) seems to fall outside of the linear trend between engine size and gas mileage. These cars have a higher mileage than you might expect. How can you explain these cars? 

```{r, echo = FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() + 
  geom_point(data = dplyr::filter(mpg, displ > 5, hwy > 20), colour = "red", size = 2.2)
```



### Aesthetics

You can add a third variable, like `class`, to a two dimensional scatterplot by mapping it to a new __aesthetic__. An aesthetic is a visual property of the objects in your plot. Aesthetics include things like the size, the shape, or the color of your points. You can display a point in different ways by changing the values of its aesthetic properties. 


We can add the `class` variable to the plot by mapping the levels of an aesthetic (like color) to the values of `class`. For example, we can color a point green if it belongs to the compact class, blue if it belongs to the midsize class, and so on. 

Let's give this a try. Fill in the blank piece of code below with `color = class`. What happens? Delete the commenting symbols (`#`) before running your code. (If you prefer British English, you can use `colour` instead of `color`.)

```{r aesthetics, exercise = TRUE, exercise.eval = FALSE}
# ggplot(data = mpg) + 
#   geom_point(mapping = aes(x = displ, y = hwy, ____________))
```

<div id="aesthetics-hint">
**Hint:** Be sure to remove all of the underlines and hashtags from the code.
</div>

```{r aesthetics-solution}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

```{r aesthetics-check}
"Great Job! You can now tell which class of car each point represents by examining the color of the point."
```

### And the answer is...

The colors reveal that many of the unusual points in `mpg` are two-seater cars. These cars don't seem like hybrids, and are, in fact, sports cars! Sports cars have large engines like SUVs and pickup trucks, but small bodies like midsize and compact cars, which improves their gas mileage. In hindsight, these cars were unlikely to be hybrids since they have large engines.

This isn't the only insight we've gleaned; you've also learned how to add new aesthetics to your graph. Let's review the process.


### Other aesthetics

To map an aesthetic to a variable, we set the name of the aesthetic equal to the name of the variable, and do this inside `mapping = aes()`. ggplot2 will automatically assign a unique level of the aesthetic (here a unique color) to each unique value of the variable. ggplot2 will also add a legend that explains which levels correspond to which values.

This insight gives us a new way to think about the mapping argument. Mappings tell ggplot2 more than which variables to put on which axes, they tell ggplot2 which variables to map to which visual properties. The x and y locations of each point are just two of the many visual properties displayed by a point.


In the above example, we mapped color to `class`, but we could have mapped size to `class` in the same way. 

Change the code below to map `size` to `class`. What happens? 

```{r size, exercise = TRUE, exercise.eval = FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

<div id="size-hint">
**Hint:** If `color` controls the _aesthetic_, what word do you suppose controls the _size_ aesthetic?
</div>

```{r size-solution}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```

```{r size-check}
"Great Job! Now the size of a point represents its class. Did you notice the warning message? ggplot2 gives us a warning here because mapping an unordered variable (class) to an ordered aesthetic (size) is not a good idea."
```

### alpha

You can also map `class` to the `alpha` aesthetic, which controls the transparency of the points. Try it below.

```{r alpha, exercise = TRUE, exercise.eval = FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

<div id="alpha-hint">
**Hint:** If `color` controls the _aesthetic_, what word do you suppose controls the _alpha_ aesthetic?
</div>

```{r alpha-solution}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))
```

```{r alpha-check}
"Great Job! If you look closely, you can spot something subtle: many locations contain multiple points stacked on top of each other (alpha is additive so multiple transparent points will appear opaque)."
```

### Shape

Let's try one more aesthetic. This time map the class of the points to `shape`, then look for the SUVs. What happened?

```{r shape, exercise = TRUE, exercise.eval = FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

<div id="shape-hint">
**Hint:** If `color` controls the _aesthetic_, what word do you suppose controls the _shape_ aesthetic?
</div>

```{r shape-solution}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```



## Exercise 1

### Task 1

Make a scatterplot of `cty` vs `hwy`.

- cty: city miles per gallon
- hwy: highway miles per gallon

```{r ex1-task1, exercise = TRUE}

```


```{r ex1-task1-solution}
ggplot(data = mpg) + 
  geom_point(aes(x = hwy, y = cty))
```


```{r ex1-task1-check}
grade_code()
```



### Task 2

Modify the code below to map `cty`, which is a continuous variable, to `color`, `size`, and `shape`. 


a) Map `cty` to color

```{r ex1-task2a, exercise = TRUE, exercise.eval = FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```


b) Map cty to size

```{r ex1-task2b, exercise = TRUE, exercise.eval = FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```


c) Map cty to shape

```{r ex1-task2c, exercise = TRUE, exercise.eval = FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```



### 

How do these aesthetics behave differently for continuous variables, like `cty`, vs. categorical variables, like `class`? 



## Histograms

Make a histogram of the `hwy` variable from the `mpg` dataset. Make the bins 2 units wide.

Hint: 
- You should not supply a $y$ aesthetic when you use `geom_histogram()`
- Use the help page for `geom_histogram` or consult the ggplot2 cheatsheet if needed.

```{r hist1, exercise=TRUE, eval=FALSE}
ggplot(data = mpg) + 
  geom_histogram(aes(x = ________),_____=______)
```


```{r hist1-solution, eval=FALSE}
ggplot(data = mpg) +
  geom_histogram(aes(x = hwy), binwidth = 2)
```

```{r hist1-check}
grade_code()
```


## Bar Charts

### Creating a bar chart

A variable is **categorical** if it can take only one of a small set of values.  You can visualize the distribution of a categorical variable with a bar chart.  

- In a bar chart, we plot the counts or frequencies of different levels of a categorical variable, by specifying the categorical variable we want to be on the x-axis, and adding a `geom_bar()` layer to the plot.
- The height of the bars reveal how many cars in the data set come from each class. 


The bar chart below visualizes the distribution of the `class` variable in the `mpg` data set, which comes in the ggplot2 package.



```{r barchart-1, exercise=TRUE}
ggplot(data = mpg) +
  geom_bar(aes(x = class)) 
```

** Remark: **

You should not supply a $y$ aesthetic when you use `geom_bar()`; ggplot2 will count how many times each $x$ value appears in the data, and then display the counts on the $y$ axis. 

### Interpretation

In histogram and bar charts, tall bars show the common values of a variable, i.e. the values that appear frequently. Shorter bars show less-common values, i.e. values that appear infrequently. Places that do not have bars reveal values that were not seen in your data. To turn this information into useful questions, look for anything unexpected:

- Which values are the most common? Why?
- Which values are rare? Why? Does that match your expectations?
- Can you see any unusual patterns? What might explain them?


### Stacked Bar Chart

A stacked bar chart adds another layer to the plot, by dividing each bar into different levels of another variable. This coloring of the bars comes from adding a second categorical variable into the `fill` argument of the `aes()` function. 

This looks something like this:

```{r barchart-2, exercise=TRUE, out.width = "100%"}
ggplot(data = mpg) +
  geom_bar(aes(x = class, fill=drv)) 
```

**Remark:**

- `drv`: the type of drive train, where f = front-wheel drive, r = rear wheel drive, 4 = 4wd

### Position Adjustment

`geom_bar()` lays out the rectangles by stacking rectangles that have the same `class` value on top of one another. You can change this behavior with a _position adjustment_.

To place rectangles that have the same `class` value beside each other, set `position = "dodge"`. Add the argument `position = "dodge"` to the `geom_bar()` below and observe the change to the bar chart.

```{r barchart-dodge, exercise=TRUE, eval=FALSE, out.width = "80%"}
ggplot(data = mpg) +
  geom_bar(aes(x = class, fill=drv), _____________) 
```

```{r barchart-dodge-solution}
ggplot(data = mpg) +
  geom_bar(aes(x = class, fill=drv), position = "dodge") 
```


## Exercise 2

### Task 1

Answer the questions using the following chart.

```{r barchart-3, echo=FALSE}
ggplot(data = mpg) +
  geom_bar(aes(x = class)) + 
  ggtitle("Classes of cars in mpg")
```

```{r ex2-task1a, echo = FALSE}
 question("What is the most common type of car in the `mpg` data set?",
           answer("2seater"),
           answer("compact"),
           answer("midsize"),
           answer("minivan"),
           answer("pickup"),
           answer("subcompact"),
           answer("suv", correct = TRUE),
           allow_retry = TRUE
  )
```

```{r ex2-task1b, echo = FALSE}

  question("What is the least common type of car in the `mpg` data set?",
           answer("2seater", correct = TRUE),
           answer("compact"),
           answer("midsize"),
           answer("minivan"),
           answer("pickup"),
           answer("subcompact"),
           answer("suv"),
           allow_retry = TRUE
  )
```



### Exercise 2

```{r echo=FALSE, out.width = "80%"}
ggplot(data = mpg) +
  geom_bar(mapping = aes(x = class, fill = class))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


Create a bar chart `class` colored by `class` to recreate the graph above.  

- Use the help page for `geom_bar` or the `ggplot` cheatsheet to choose a "color" aesthetic for class.
- The text labels in the x-axis in the plot may be overlapping with each others. We should add an extra layer to the above plot such that the x-Axis labels are rotated by 45 degrees with right-justified for horizontal justification.

```
    + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r ex2-task2, exercise=TRUE, out.width = "80%", eval=FALSE}
ggplot(_____________) +
  ______________________ +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r ex2-task2-solution, out.width = "80%"}
ggplot(data = mpg) +
  geom_bar(mapping = aes(x = class, fill = class))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Facets


Faceting breaks the data into subsets based on the levels of a categorical variable and then constructs a subplot for each level.

ggplot2 provides two functions for facetting: `facet_grid()` and `facet_wrap()`. 

### `facet_wrap()`

`facet_wrap()` provides a  way to facet a plot on a _single_ variable. It will split the plot into subplots and then reorganize the subplots into multiple rows so that each plot has a more or less square aspect ratio. 

To add a faceting layer we use the `facet_wrap()` function. Inside the `facet_wrap()` function we add a tilde (`~`) and then the name of the variable we want to facet by (e.g. `~ class`). This can be read as creating the above plot "broken down by class".

**Example:**

```{r facet-wrap1, exercise=TRUE}
ggplot(mpg) + geom_point(aes(x = displ, y = hwy)) +
  facet_wrap(~ class)

```


### facet_grid()

`facet_grid()` divides the plot into a grid of subplots based on the values of one or two facetting variables. To use it, add `facet_grid()` to the end of your plot call. 

** Example:**

```{r facet-grid1, exercise=TRUE}
q <- ggplot(mpg) + geom_point(aes(x = displ, y = hwy))
q + facet_grid(drv ~ cyl)
q + facet_wrap(~ class)

```

As you saw in the code examples, you use `facet_grid()` by passing it a formula, the names of two variables connected by a `~`.  

- `facet_grid()` will split the plot into facets vertically by the values of the first variable: each facet will contain the observations that have a common value of the variable. 
- `facet_grid()` will split the plot horizontally by values of the second variable. The result is a grid of facets, where each specific subplot shows a specific combination of values. 
- If you do not wish to split on the vertical or horizontal dimension, pass `facet_grid()` a `.` instead of a variable name as a place holder.


** Example:**

```{r facet-grid2, exercise=TRUE}
q <- ggplot(mpg) + geom_point(aes(x = displ, y = hwy))
q + facet_grid(. ~ cyl)
q + facet_grid(drv ~ .)

```


## Boxplot

### Creating a boxplort
Boxplot allows you to visualize the distribution of continuous variables.

Here is a boxplot for showing the distribution of highway miles per gallon (`hwy`).

```{r echo = TRUE, out.width = "100%"}
ggplot(data = mpg) + 
  geom_boxplot( aes( y = hwy))
```

### Relationship between one continuous and one categorical variable

You can visualize the relationship between one continuous and one categorical variable by using a boxplot. 
For instance, to create a box plot to compare the fuel efficiencies of different types of cars, modify the code below to add a mapping of the y-xis to the  `hwy` attribute.


```{r boxplot1, exercise = TRUE, eval=FALSE}
ggplot(data = mpg) + 
  geom_boxplot(aes(x = ________ , y = hwy))
```


```{r boxplot1-solution}
ggplot(data = mpg) + 
  geom_boxplot( aes(x = class, y = hwy))
```

## Exercise 3

```{r ex3-result, echo=FALSE, out.width = "100%"}
ggplot(data = mpg) + 
  geom_boxplot(mapping = aes(x = class, y = hwy, fill=class)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Recreate the chart above. 
- The text labels in the x-axis in the plot may be overlapping with each others. We should add an extra layer to the plot such that the x-Axis labels are rotated by 45 degrees with right-justified for horizontal justification. (Refer to exercise 2)

```{r ex3, exercise = TRUE, eval=FALSE, out.width = "100%", exercise.reveal_solution=FALSE}


```



```{r ex3-solution}
ggplot(data = mpg) + 
  geom_boxplot(mapping = aes(x = class, y = hwy, fill=class))


```

