---
title: "Introduction to linear regression"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE,message=FALSE}
library(learnr)
library(learnrhash)
library(tidyverse)
library(gradethis)
library(openintro)
library(infer)
tutorial_options(
  # use gradethis for checking
  exercise.checker = gradethis::grade_learnr
  )
knitr::opts_chunk$set(echo = FALSE)


```
## Getting Started

The Human Freedom Index is a report that attempts to summarize the idea of "freedom"
through a bunch of different variables for many countries around the globe.
It serves as a rough objective measure for the relationships between the different
types of freedom - whether it's political, religious, economical or personal freedom -
and other social and economic circumstances. The Human Freedom Index is an annually
co-published report by the Cato Institute, the Fraser Institute, and the Liberales
Institut at the Friedrich Naumann Foundation for Freedom.

In this lab, you'll be analyzing data from Human Freedom Index reports from 2008-2016.
Your aim will be to summarize a few of the relationships within the data both 
graphically and numerically in order to find which variables can help tell a story
about freedom.

### Load packages

In this lab, you will explore and visualize the data using the **tidyverse** suite of 
packages. The data can be found in the companion package for OpenIntro resources, **openintro**.

Let's load the packages. 

```{r load-packages, exercise = T}
library(tidyverse)
library(openintro)
```


### The data

The data we're working with is in the openintro package and it's called `hfi`, 
short for Human Freedom Index.

```{r}
data(hfi)
```


Plot the relationship between `pf_expression_control` (political pressures and controls on media content) and the personal freedom score, `pf_score` with `pf_expression_control` as the explanatory/predictor variable. Does the relationship look linear? 

```{r nvolifwefsdohnqx, exercise = TRUE}
ggplot(hfi, aes(x = ___, y = ___))+geom_point()
```


```{r nvolifwefsdohnqx-solution}
ggplot(hfi, aes(x = pf_expression_control, y = pf_score))+geom_point()

```

```{r nvolifwefsdohnqx-check}
# check code
gradethis::grade_code()
```

If the relationship looks linear, we can quantify the strength of the
relationship with the correlation coefficient.

```{r cor, exercise = T}
hfi %>%
  summarise(cor(pf_expression_control, pf_score, use = "complete.obs"))
```

Here, we set the `use` argument to "complete.obs" since there are some observations of NA and we don't want those to count towards the calculation of the correlation.

## The linear model

You can use the `lm` function in R to fit the linear model (a.k.a. 
regression line).

- The first argument in the function `lm` is a formula that takes the form 
`y ~ x`. 
  - In this example,  we want to make a linear model of `pf_score` as a function of `pf_expression_control`. 
- The second argument specifies that R should look in the `hfi` data frame to find the two variables. 

**Note:** `lm` can handle missing values, so we can use the full data set here.



```{r m1, exercise = T}
m1 <- lm(pf_score ~ pf_expression_control, data = hfi)
```

The output of `lm` is an object that contains all of the information we need 
about the linear model that was just fit. We can access this information using 
the summary function.

```{r summary-m1-setup}
m1 <- lm(pf_score ~ pf_expression_control, data = hfi)
```

```{r summary-m1, exercise = TRUE}
summary(m1)
```

Let's consider this output piece by piece. 

- The formula used to describe the model is shown at the top. 
- After the formula you find the five-number 
summary of the residuals. 
- The "Coefficients" table shown next is key; its first 
column displays the linear model's y-intercept and the coefficient of `pf_expression_control`.

With this table, we can write down the least squares regression line for the 
linear model:

\[
  \hat{y} = 4.61707 + 0.49143 \times pf\_expression\_control
\]

One last piece of information we will discuss from the summary output is the 
Multiple R-squared, or more simply, $R^2$. The $R^2$ value represents the 
proportion of variability in the response variable that is explained by the 
explanatory variable. For this model, 63.42% of the variability in personal freedom score is explained by political pressures and controls on media content.

Fit a new model that uses `pf_expression_control` to predict the total human freedom score, `hf_score`. 

```{r adryxdkevttxzzno, exercise = TRUE}
new_mod <- lm(___ ~ ___, data = ___)
summary(___)
```

```{r adryxdkevttxzzno-solution}
new_mod <- lm(hf_score ~pf_expression_control, data = hfi)
summary(new_mod)
```

```{r adryxdkevttxzzno-check}
# check code
gradethis::grade_code()
```

Using the estimates from the R output, write the equation of the regression line. 

```{r regline}
question("Using the estimates from the R output, what is the equation of the regression line?",
    answer("hf_score_hat = 5.15 + 0.35 * pf_expression_control", correct = T),
    answer("hf_score = 5.15 + 0.35 * pf_expression_control"),
    answer("hf_score_hat = 0.35 + 5.15 * pf_expression_control"),
    answer("pf_expression_control_hat = 5.15 + 0.35 * hf_score"),
    answer("pf_expression_control_hat = 0.35 + 5.15 * hf_score"),
    answer("pf_expression_control = 0.35 + 5.15 * hf_score"),
    allow_retry = TRUE,
    random_answer_order = F
  )
```


## Prediction and prediction errors

Let's create a scatterplot with the least squares line for `m1` laid on top.

```{r reg-with-line, exercise = T}
ggplot(data = hfi, aes(x = pf_expression_control, y = pf_score)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

Here, we are literally adding a layer on top of our plot. `geom_smooth` creates
the line by fitting a linear model. It can also show us the standard error `se`
associated with our line, but we'll suppress that for now (and talk about what it means later).

This line can be used to predict $y$ at any value of $x$. When predictions are 
made for values of $x$ that are beyond the range of the observed data, it is 
referred to as *extrapolation* and is not usually recommended. However, 
predictions made within the range of the data are more reliable. They're also 
used to compute the residuals.

Remember the regression line is:
\[
  \hat{y} = 4.61707 + 0.49143 \times pf\_expression\_control
\]

```{r predval}
question("What is the predicted value of a country's personal freedom score for one with a 6.7 rating for `pf_expression_control`? ",
    answer("4.62"),
    answer("7.91", correct = T),
    answer("31.43"),
    allow_retry = TRUE,
    random_answer_order = F
  )
```

```{r resid}
question("If the true personal freedom score for this observation is 8.9, what is the residual for this prediction?  ",
    answer("0.99", correct = T),
    answer("-0.99"),
    answer("0"),
    allow_retry = TRUE,
    random_answer_order = F
  )
```


