---
title: "logistic regression"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
renv::dependencies()
```


## Ex

```{r}
#data from https://en.wikipedia.org/wiki/Logistic_regression
# ref: https://rpubs.com/johnakwei/167443
(hours=c(0.50,0.75,1.00,1.25,1.50,1.75,1.75,2.00,2.25,2.50,2.75,3.00,3.25,3.50,4.00,4.25,4.50,4.75,5.00,5.50))
(pass = c(0,0,0,0,0,0,1,0,1,0,1,0,1,0,1,1,1,1,1,1))
(mydata<-data.frame(hours, pass))


```

```{r}
ggplot() + 
   geom_point(data=mydata, aes(x=hours, y=pass, color="admit")) 

```


```{r}
model <- glm(pass ~ hours,family=binomial(link='logit'),data=mydata)
model$coefficients
```


Probability of passing exam =1/(1+exp(-(-4.0777+1.5046* Hours)))

For example, for a student who studies 2 hours, entering the value Hours =2 in the equation gives the estimated probability of passing the exam of p=0.26:

StudentHours <- 2
ProbabilityOfPassingExam <- 1/(1+exp(-(-4.0777+1.5046*StudentHours)))
ProbabilityOfPassingExam


Similarly, for a student who studies 4 hours, the estimated probability of passing the exam is p=0.87:
Probability of passing exam =1/(1+exp(-(-4.0777+1.5046*4))) = 0.87.

StudentHours <- 4
ProbabilityOfPassingExam <- 1/(1+exp(-(-4.0777+1.5046*StudentHours)))
ProbabilityOfPassingExam



```{r}
x = seq(0,6,0.25)
y = 1/(1+exp(-(-4.0777+1.5046* x)))
d = data.frame(x,y)
ggplot()+
  geom_point(data=d, aes(x=x,y=y))+
  geom_vline(aes(xintercept = 2.710156), linetype="dotted")
```

let $z=-4.0777+1.5046* x$.

```{r}
library(ggplot2)
ggplot()+
  scale_x_continuous(name="x", limits=c(-10,40)) +
  scale_y_continuous(name="y", limits=c(-10,50)) +
 # scale_linetype(name="s") +
  geom_abline(slope=1.5046, intercept=-4.0777) +
  geom_line() +
  coord_fixed(ratio=0.5)
```





```{r}
z = seq(-5,6,0.25)
y = 1/(1+exp(-(z)))
d = data.frame(z,y)
ggplot()+
  geom_point(data=d, aes(x=z,y=y))
```

$-4.0777+1.5046* x = 0 $

```{r}
x = 4.0777/1.5046
x
```
When # of hours of study = 2.710156, prob(pass)=0.5







```{r}
Hours <- c(0.50, 0.75, 1.00, 1.25, 1.50, 1.75, 1.75, 2.00, 2.25,
           2.50, 2.75, 3.00, 3.25, 3.50,    4.00,   4.25,   4.50,   4.75,
           5.00, 5.50)
Pass    <- c(0, 0, 0, 0, 0, 0, 1,   0, 1, 0, 1, 0, 1, 0, 1, 1, 1,   1, 1, 1)
HrsStudying <- data.frame(Hours, Pass)
ggplot(HrsStudying, aes(Hours, Pass)) +
  geom_point() +
  #geom_smooth(method='glm', family="binomial", se=FALSE) +
   geom_smooth(method = 'glm', se=FALSE, method.args= list(family="binomial"))+
  labs (x="Hours Studying", y="Probability of Passing Exam",
        title="Probability of Passing Exam vs Hours Studying")
```



```{r}
mymodel <- glm(pass ~ hours, #+SibSp+Age,        # Formula
                    data= mydata,    # Data set
                    family="binomial")      # family="binomial" for binary logistic

summary(mymodel)                      # Check model summary
```

The equation is

$pass = $

```{r}
mymodel <- glm(pass ~  hours , data = mydata, family = "binomial")
summary(mylogit)

```




## Example

. A researcher is interested in how variables, such as GRE (Graduate Record Exam scores), GPA (grade point average) and prestige of the undergraduate institution, effect admission into graduate school. The response variable, admit/don’t admit, is a binary variable.


```{r}
mydata <- read.csv("https://stats.idre.ucla.edu/stat/data/binary.csv")

## view the first few rows of the data
str(mydata)
head(mydata)
```


- This dataset has a binary response (outcome, dependent) variable called admit. 
- There are three predictor variables: gre, gpa and rank.
  - the variables gre and gpa as continuous. The variable rank takes on the values 1 through 4. 
  - Institutions with a rank of 1 have the highest prestige, while those with a rank of 4 have the lowest. 


```{r}
## two-way contingency table of categorical outcome and predictors we want
## to make sure there are not 0 cells
xtabs(~admit + rank, data = mydata)
```


```{r}
mydata$rank <- factor(mydata$rank)
mylogit <- glm(admit ~  gpa , data = mydata, family = "binomial")
summary(mylogit)
```

```{r}
df<-data.frame(gpa=seq(-1, 5, by = 0.1))
df

prob_admit <- predict(mylogit,              # Model to use
                       newdata=df,      # Data to use for predictions
                       type="response")            # Return predicted probabilities
(mydata2<-cbind(prob_admit,df))
```


```{r}
ggplot() + 
   geom_point(data=mydata, aes(x=gpa, y=admit, color="admit")) 
  #+ geom_point(data=mydata2, aes(x=gpa, y=prob_admit, color="predicted"))
  
```


## Titanic

```{r}
install.packages("titanic")
library(titanic)
head(titanic_train)
titanic_train
```
```{r}
str(titanic_train)
```

```{r}
titanic_train$Name <- as.character(titanic_train$Name)    # Convert name to character

titanic_train$Pclass <- ordered(titanic_train$Pclass,     # Convert to ordered factor
                                levels=c("3","2","1")) 
titanic_train$Survived<-factor(titanic_train$Survived)
titanic_train$Name <- NULL  # Remove name column
titanic_train$Ticket <- NULL  # Remove ticket column
titanic_train$Cabin <- NULL  # Remove Cabin column
str(titanic_train)
```

```{r}
head(titanic_train)
```



```{r}
summary(titanic_test)
```

```{r}
?titanic
```



```{r}
options(scipen=999)   # avoid using scientific notation
titanic_model <- glm(Survived ~ ., #+SibSp+Age,        # Formula
                    data= titanic_train,    # Data set
                    family="binomial")      # family="binomial" for binary logistic

summary(titanic_model)                      # Check model summary
```

```{r}
train_preds <- predict(titanic_model,              # Model to use
                       newdata=titanic_train,      # Data to use for predictions
                       type="response")            # Return predicted probabilities

table(train_preds, titanic_train$Age)
```



```{r}
mydata<-titanic_train[1-10,]
survived_prob <- predict(titanic_model,              # Model to use
                       newdata=mydata,      # Data to use for predictions
                       type="response")            # Return predicted probabilities
#mydata$predicted <- train_preds
mydata<-cbind(predicted_survived,mydata)
mydata$Predicted_Survived = ifelse(mydata$predicted_survived >= 0.5, 1, 0)

#colnames(mydata)
mydata[,c("predicted_survived","Survived","Predicted_Survived","Age")]
#table(train_preds, titanic_train$Sex)
```

```{r}
library(ggplot2)
ggplot(mydata, aes(x=Age, y=predicted_survived)) +
  geom_point(size=2, shape=23)
```



## Draw a line

```{r}
library(ggplot2)
data(iris)
str(iris)
x=1:5
ggplot(iris, aes(x=Sepal.Length, y=Sepal.Width, col = Species)) +
  geom_point(alpha = 1) +
  stat_function(fun=function(x) {x-3}, color = "grey", size = 1) 
```

## Diabetest

```{r}
#ref: https://rpubs.com/Mayank7j_2020/PimaIndiansDiabetesData
#install.packages("mlbench")
library(mlbench) #data sets from the UCI repository.
data("PimaIndiansDiabetes")
# rename dataset to keep code below generic
dataset <- PimaIndiansDiabetes
str(dataset)
```



```{r}
model <- glm(diabetes ~ ., ,        # Formula
                    data= PimaIndiansDiabetes,    # Data set
                    family="binomial")      # family="binomial" for binary logistic

summary(model)  
```
```{r}
data<-dataset[, -which(names(dataset) == "diabetes")]  #remove the diabetes column
data
```

```{r}
library("ggplot2")
ggplot(dataset, aes(x=glucose, fill=diabetes)) +
  geom_bar( position='dodge', stat = "bin",  binwidth = 30) 
```

```{r}
library("ggplot2")
ggplot(dataset, aes(x=pressure, fill=diabetes)) +
  geom_bar( position='dodge', stat = "bin",  binwidth = 30) 
```
```{r}
library("ggplot2")
ggplot(dataset, aes(x=age, fill=diabetes)) +
  geom_bar( position='dodge', stat = "bin",  binwidth = 10) 
```



```{r}
#install.packages("corrplot")
library(corrplot)
cor_data <- cor(data)
cor_data
corrplot(cor_data,method = "number")
```


```{r}
library("ggplot2")
ggplot(dataset, aes(x=age, y=glucose, col=diabetes)) +
  geom_point( ) +
  stat_function(fun=function(x) {165.875 +  -0.6951521*x}, color = "black", size = 0.5) 

```

```{r}
model <- glm(diabetes ~ glucose+age, ,        # Formula
                    data= PimaIndiansDiabetes,    # Data set
                    family="binomial")      # family="binomial" for binary logistic

summary(model)  
```

-5.912449 + 0.035644*glucose+0.024778*age=0
-5.912449 + 0.024778*age= -0.035644*glucose 
165.875 +  -0.6951521*age= glucose 

```{r}
 0.024778/-0.035644
```
## Iris

Convert setosa and versicolor species to others.

```{r}
unique(iris$Species)
library(tidyverse)
data(iris)
mydata <- iris
mydata$Species <- as.character(mydata$Species)
mydata$Species[mydata$Species == 'setosa'] <- 'others'
mydata$Species[mydata$Species == 'versicolor'] <- 'others'
unique(mydata$Species)
```

```{r}
unique(mydata$Species)
```


```{r}
library("ggplot2")

ggplot(mydata, aes(y=Petal.Length, x=Sepal.Length, col=Species)) +
  geom_point( ) +
  scale_colour_manual(values = c("tomato","blue")) 

```


```{r}
data(iris)
iris
mydata=iris[1:100,]
unique(mydata$Species)

model <- glm(Species ~ Sepal.Length+Petal.Length,         # Formula
                    data= mydata,    # Data set
                    family="binomial")      # family="binomial" for binary logistic

summary(model)  

```



```{r}

library(tidyverse)
df <- data.frame(expand.grid(1:10,1:10))
x<-1:5
y<-1:5
df <- data.frame(x=x,y=y)
#df %>%
#  mutate(z, Var1+Var2)
df %>%
  mutate(z=x+y)

#ggplot(mydata, aes(x=Sepal.Length, y=Petal.Length,  z=Sepal.Width, col=Species))  + 
##    geom_point()+
#    geom_contour_filled() 



```

z=65.84 - 37.56*Sepal.Length + 49.05*Petal.Length


```{r}
#install.packages("metR")
library(tidyverse)
library(metR)
 <- data.frame(expand.grid(1:10,1:10))
df
df<-df %>%
  mutate(Var3=1/(1+exp(-1*(Var1*(-37.56)+Var2*(49.05)+65.84))))
df

```



```{r}
library(plotly)
Sepal.Length = 1:10
Petal.Length = 1:10

#z=65.84 - 37.56*Sepal.Length + 49.05*Petal.Length

my_function <- function(x,y) {
    final_value = 1/(1+exp(-1*(Sepal.Length*(-37.56)+y*(49.05)+65.84)))
}

z <- outer(x, y, my_function)
print(z)
plot_ly(x = ~Sepal.Length, y = ~Petal.Length, z = ~z, opacity=0.6,colorscale ='Blues')  %>% 
        add_surface() %>%
    layout(
    xaxis = list(range = c(100, 1), autorange = F, autorange="reversed"), 
    yaxis = list(range = c(2, 5)))
```

```{r}
library(plotly)

s <- seq(1, 8)
plot_ly(x = s, y = s) %>%
  add_trace(y = rev(s)) %>%
  layout(
    xaxis = list(range = c(100, 1), autorange = F, autorange="reversed"), 
    yaxis = list(range = c(2, 5)))
```



```{r}
fig2 <-  plot_ly(data = iris ,x =  ~Sepal.Length, y = ~Sepal.Width, color = ~Species, type = 'scatter', mode = 'markers')%>%
  layout(title = 'Manually Specified Labels', plot_bgcolor = "#e5ecf6", xaxis = list(title = 'Sepal Length (cm)'), 
         yaxis = list(title = 'Sepal Width (cm)'), legend = list(title=list(text='<b> Species of Iris </b>')))
fig2
```


```{r}
ggplot()  + 
  geom_contour_filled(data=df, aes(x=Var1, y=Var2,  z=Var3)) + 
  scale_fill_brewer(palette = "Blues")
```




```{r}
mydata=iris[1:100,]
ggplot()  + 
  geom_contour_filled(data=df, aes(x=Var1, y=Var2,  z=Var3)) + 
  scale_fill_brewer(palette = "Blues")+
  geom_point(data=mydata, aes(y=Petal.Length, x=Sepal.Length, col=Species)) +
  scale_colour_manual(values = c("tomato","blue")) +
  labs(title="Scatterplot",
        x ="Sepal Length", y = "Petal Length")
```



## References

- https://stats.oarc.ucla.edu/r/dae/logit-regression/
- https://www.kaggle.com/code/hamelg/intro-to-r-part-28-logistic-regression/notebook



```{r}
library(modeldata)
data(attrition)
attrition
```

