---
title: "Titanic Dataset: Exploratory Data Analysis"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
tutorial:
  id: "demo-types"
  version: 1.0
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
library(gradethis)
library(titanic)
library(tidyverse)

check_result <-function(){
    grade_this({
    if (.solution == .result)
      pass()
    fail()
  }
  )
}
#convert variable to factors
titanic_train$Survived<-factor(titanic_train$Survived)
titanic_train$Pclass<-factor(titanic_train$Pclass)

#remove unnecessary variables.
rm(titanic_test, titanic_gender_class_model, titanic_gender_class_model)
```


## Loading the libararies and dataset
###
Titanic is an R package containing data sets providing information on the fate of passengers on the fatal maiden voyage of the ocean liner "Titanic".

First, we  load the dataset and libraries. In this tutorial, we will make use of the data frame `titanic_train`.

```{r echo=TRUE}
library(titanic)
library(tidyverse)
```


### 

Here is the preview of the data set.

```{r echo=TRUE}
titanic_train
```


###

Let's examine the structure of the data frame.

```{r glimpse, exercise=TRUE}
glimpse(titanic_train)
```

###

Here is the description of the variables:

- **survival**: Survival (0 = No; 1 = Yes)
- **Pclass**: Passenger Class (1 = 1st; 2 = 2nd; 3 = 3rd)
- **name**: Name of passenger
- **sex**: Gender of passenger
- **age**: Age of passenger
- **sibsp**: Number of Siblings/Spouses Aboard
- **parch**: Number of Parents/Children Aboard
- **ticket**: Ticket Number
- **fare**: Passenger Fare (British pound)
- **cabin**: Cabin
- **embarked**: Port of Embarkation (C = Cherbourg; Q = Queenstown; S = Southampton)

## Preprocessing 

**Exercise**:
Convert the variables `Survived` and `Pclass` of the `titanic_train` data frame to factor variables.

```{r to-factor-setup}
#convert the variables to int to simulate the original dataset
titanic_train$Survived<-as.integer(titanic_train$Survived)
titanic_train$Pclass<-as.integer(titanic_train$Pclass)
```


```{r to-factor, exercise=TRUE, eval=FALSE}
titanic_train$Survived<-________
titanic_train$Pclass<-_________

#Check that the data type of the variables are factor variables.
glimpse(select(titanic_train, Survived, Pclass))

```

```{r to-factor-solution}
titanic_train$Survived<-factor(titanic_train$Survived)
titanic_train$Pclass<-factor(titanic_train$Pclass)
```


```{r to-factor-check}
grade_this({
  if (exists("titanic_train", envir=.envir_result))
    titanic_train_result <- get("titanic_train", envir = .envir_result)
    
  glimpse(titanic_train_result)
  if (is.factor(titanic_train_result$Survived) && is.factor(titanic_train_result$Pclass)){
    pass()
  }
  fail("Please convert the variables to factor variables.")
}
)
```

## Exploring the dataset


### 

Examine the few rows of data in the training set using the `head` function.

```{r head-solution}
head(titanic_train)
```


```{r head, exercise=TRUE}

```



Use the `summary()` function to view the summary statistics on the dataset.
```{r summary-solution}
summary(titanic_train)
```



```{r summary, exercise=TRUE}

```


**Question:** Which variable has missing values (NA)?

### Contingency Table

The `table` function counts the occurrence of each value in a vector. What does the following command output?


```{r gender, exercise=TRUE}
table(titanic_train$Sex) 
```

###

**Exercise:**
In the training set, how many male passengers are there?

```{r male-count, echo=FALSE}
question_text(
  "Answer: ",
  answer("577", correct = TRUE),
  incorrect = "It's not correct.",
  #try_again_button = "Modify your answer",
  allow_retry = TRUE
)
```

###

We can send the output of the `table` command to  `prop.table()` to compute the proportion.

```{r prop_gender, exercise=TRUE}
t<-table(titanic_train$Sex) 
prop.table(t)
```

###

**Exercise:**
Complete the code to find out the proportion of passenger survived using the `prob.table` function.

```{r prop_survived-solution}
t<-table(titanic_train$Survived) 
prop.table(t)
```

```{r prop_survived, exercise=TRUE}

```

```{r prop_survived-check}
check_result()
```


###
<div id="prop-survivor-ex">
**Exercise:**
What is the proportion of survivors (rounded to 2 decimal places)? 
</div>


```{r details, echo=FALSE}
question_text(
  "Answer: ",
  answer("0.62", correct = TRUE),
  incorrect = "It's not correct.",
  #try_again_button = "Modify your answer",
  allow_retry = TRUE
)
```



## Charts

### Survival by Gender

Create a bar chart to show the distribution of male and female passengers.

```{r barchart-gender-solution}
ggplot(titanic_train,aes(x=Sex))+
  geom_bar() 
```


```{r barchart-gender, exercise=TRUE}


```

---

###

To understand if male or female is more likely to survive, compute the proportion of survivors grouped by Gender by passing the vectors `titanic_train$Sex` and `titanic_train$Survived` below.

```{r prop_sex_survived, exercise=TRUE, eval=FALSE}
t<-table(_____, ______) 
prop.table(t)
```

---

###


Complete the code to replicate the bar chart below.

```{r echo=FALSE}
ggplot(titanic_train,aes(x=Sex,fill=Survived))+
  geom_bar()
```


```{r plot_survived_sex-solution}
ggplot(titanic_train,aes(x=Sex,fill=Survived))+
  geom_bar()
```


```{r plot_survived_sex, exercise=TRUE}

```


---

Is Male or Female passenger more likely to survive? 

```{r survive_prob, echo=FALSE}
question("Which of the following statements is TRUE related to chance of survival of female and male?",
  answer("Male passengers have higher probability of survival than female", message = "Nope, not quite!"),
  answer("Female passengers have higher probability of survival than male", correct = TRUE),
  answer("Male and female passengers have equal probability of survival",  message = "Try again!"), 
  allow_retry = TRUE
)
```



###  Survival by passenger class

**Exercise:** Plot a bar chart to the number of passengers for each passenger class.

```{r Pclass-solution}
ggplot(titanic_train,aes(x=Pclass))+
  geom_bar() 
```

```{r Pclass, exercise=TRUE}

```

---

###

To understand if the passenger class may affect the survival probability, complete the code below to replicate the following bar chart.

```{r echo=FALSE}
ggplot(titanic_train,aes(x=Pclass, fill=Survived))+
  geom_bar() 
```



```{r barchart_PClass-solution}
ggplot(titanic_train,aes(x=Pclass, fill=Survived))+
  geom_bar() 
```

```{r barchart_PClass, exercise=TRUE}

```

---


```{r survive_prob_class, echo=FALSE}
question("Which of the following passenger class has the highest chance of survival?",
  answer("Class 2", message = "Nope, not quite!"),
  answer("Class 3",  message = "Try again!"), 
  answer("Class 1", correct = TRUE),
  allow_retry = TRUE
)
```


## Missing values

### Exploring the passengers' age

**Exercise:** Plot the histogram show the age distribution of passengers (with bindwidth of 10).

```{r histogram_age, exercise=TRUE}
ggplot(titanic_train, aes(x=Age)) +
   geom_histogram(binwidth = 10)
```


Compute the average age of the passenger by passing the vector containing the passenger's age from the `titanic_train` data frame to the `mean` function. 

```{r mean-age-solution}
mean(titanic_train$Age)
```

```{r mean-age, exercise=TRUE}


```



###

As there are missing values in the age attributes, you cannot directly compute the average of the `age` attribute. 
We can use the option `na.rm = TRUE` of the `mean` function to ignore the missing values (NA).

```{r mean-age2-solution}
mean(titanic_train$Age, na.rm = TRUE)
```


```{r mean-age2, exercise=TRUE}
mean(titanic_train$Age, ________)
```

### Handling missing values

Missing values are common in data and need to be carefully handled in the initial stages of analysis. To examine the number of rows of missing values for each attribute, you may execute

```{r missing, echo=TRUE}
colSums(is.na(titanic_train))
```

###

We can use the `visdat` package to visualize the missing values. `vis_dat` visualises the whole dataframe at once, and provides information about the class of the data input into R, as well as whether the data is missing or not. Provide the data frame `titanic_train` as the parameter for the `vis_data` function below.


```{r visualize_NA, exercise=TRUE}
library(visdat)
vis_dat(titanic_train)
```

### 

One way to to handle missing values is to replace them with the average age of passengers. Complete the code below to replace the missing values in the `Age` attribute of the `titanic_train` dataset.


```{r handle_NA, exercise=TRUE, eval=FALSE}
#handle missing values
avg<-mean(_______, _______)
titanic_train$Age[is.na(titanic_train$Age)] = avg

#Verify that there are no more missing values.
vis_dat(___________)
```


```{r handle_NA-solution}
avg<-mean(titanic_train$Age, na.rm = TRUE)
titanic_train$Age[is.na(titanic_train$Age)] = avg
```


## Discretize variables

The attribute `Age` is a continuous variable which makes drawing proportion tables almost useless as there may only be one or two passengers for each age! Let’s create a factor variable `IsAdult` to indicate whether the passenger is a child (below the age of 18) or adult (aged 18 or above). 

```{r discretize_age-solution}
#Create the isAdult variable
titanic_train$isAdult[titanic_train$Age >= 18] <- "Adult"
titanic_train$isAdult[titanic_train$Age < 18] <- "Child"

#convert the isAdult attribute to factor
titanic_train$isAdult <-factor(titanic_train$isAdult)
```

```{r discretize_age, exercise=TRUE, eval=FALSE}
#Create the isAdult variable
titanic_train$isAdult[____________] <- "Adult"
titanic_train$isAdult[____________] <- "Child"

#convert the isAdult attribute of titanic_train data frame to factor


```

###

By using the `titanic_train$isAdult` attribute created in the previous step, compute the number of of children and adults in the dataset using the `table` function.

```{r isAdult-prob, exercise=TRUE, exercise.setup="discretize_age-solution"}
table(________)
```


```{r isAdult-prob-solution}
table(titanic_train$isAdult)
```

###

Is Child or Adult more likely to survive? Let's create a bar chart to show the number of survived/killed passenger for children and adults.


Complete the code to replicate the following chart.

```{r echo=FALSE}
#preparation
titanic_train$isAdult[titanic_train$Age >= 18] <- "Adult"
titanic_train$isAdult[titanic_train$Age < 18] <- "Child"
titanic_train$isAdult <-factor(titanic_train$isAdult)

#plot chart
ggplot(titanic_train,aes(x=isAdult,fill=Survived))+
  geom_bar(position="dodge") 

```

```{r barchart_isAdult-solution}
ggplot(titanic_train,aes(x=isAdult,fill=Survived))+
  geom_bar(position="dodge") 

```


```{r barchart_isAdult, exercise=TRUE, exercise.setup="discretize_age-solution"}


```



```{r adult_child_survival, echo=FALSE}
question("Which of the following statements is TRUE related to survival of Adults and Children?",
  answer("Adults had higher probability of survival than Children", message = "Nope, not quite!"),
  answer("Children had higher probability of survival than Adults", correct = TRUE),
  answer("Children and Adults were equally likely to survive",  message = "Try again!"), 
  allow_retry = TRUE
)
```

## References

- https://cran.r-project.org/web/packages/explore/vignettes/explore_titanic.html
- https://rolkra.github.io/explore-count-data/
- https://cran.r-project.org/web/packages/DataExplorer/vignettes/dataexplorer-intro.html
- https://www.youtube.com/watch?v=Zx2TguRHrJE&list=PLLSmf_EmjdvTiXrtivMJA63SR7yhxgTic&index=5&t=792s
- https://www.youtube.com/watch?v=GSk-EEu1zkA&list=PLLSmf_EmjdvTiXrtivMJA63SR7yhxgTic&index=3&t=573s
- https://trevorstephens.com/kaggle-titanic-tutorial/getting-started-with-r/
- https://r.smartana.org/
